Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-26 bbbq_article.tex
Add number of SNPs in NCBI database

The number of variations
was limited to the first 250 variations per gene,
resulting in 282250 variations. 
A variation needs not always be a SNP,
as dbSNP also catalogs other DNA alterations, such as, among others, insertions,
deletions and tandem repeats.
We select only the variations that result in a SNP for
a single amino acid substitution.

@@

In the final set of experiments, 
we addressed the question whether TMHs 
would be more evolutionary conserved than soluble proteins regions 
by comparing the occurrences of SNPs in the genome coding 
for TMHs and soluble protein regions. 
We obtained 1,129 gene names associated with the phrase 'membrane protein'.
These genes are linked to 4,811 protein isoforms, 
of which 2,568 are predicted to be transmembrane proteins (TMPs),
where the others are predicted to be membrane-associated proteins.
We obtained 38,882 SNPs that resulted in an
amino acids substitution, of which 21,738 were located 
in predicted transmembrane proteins.
Per protein, we calculated two percentages: 
(1) the percentage of the gene coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without membrane-associated proteins,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs and soluble protein domains.
Our analysis shows that SNPs occur more often in soluble domains
than in TMHs.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

To statistically test whether SNPs occur just as likely in soluble as
in transmembrane protein domains, we used a Poisson binomial
distribution.
We calculated the number of SNPs that are expected to be
found in TMHs by chance, $E(n_{SNP\_TMH})$, when including
and excluding soluble proteins (with symbols
$E(n_{SNP\_TMH}^{+})$ and $E(n_{SNP\_TMH}^{-})$ respectively).
The number of SNPs that are expected to be
found in TMHs by chance when including soluble proteins,
$E(n_{SNP\_TMH}^{+})$ 
equals the number of (all) SNPs, $n_{SNP}^{+} = 38,882$, multiplied by the fraction
of all proteins that are TMH, $f_{TMH}^{+} = 9.96\%$,
resulting in 
$E(n_{SNP\_TMH}^{+}) = n_{SNP}^{+} \cdot f_{TMH}^{+} \approx 3873$
The number of SNPs that are expected to be
found in TMHs by chance when excluding soluble proteins,
$E(n_{SNP\_TMH}^{-})$ 
equals the number of SNPs in TMHs, $n_{SNP}^{-} = 21,738$, 
multiplied by the fraction of the TMPs that are TMH, $f_{TMH}^{-} = 18.66\%$,
$E(n_{SNP\_TMH}^{-}) = n_{SNP}^{-} \cdot f_{TMH}^{-} \approx 4056$.
The discrepancy between the two expectations
is caused by the distribution of SNPs over the soluble and
transmembrane proteins: 
there are 38,882 SNPs found in 4811 proteins (i.e. $\approx 8.08$ SNP per protein),
of which 21,738 SNPs in 2568 TMPs (i.e. $\approx 8.46$
SNP per TMP).

@@

We first calculated the expected number of SNPs
found in TMHs, which is 61,457 SNPs \richel{GvdB: I don’t completely follow this. Based on what?} and multiplied this number by 6.8\% 
of all proteins being a TMH, resulting in the null hypothesis
of 4203 SNPs occurring in TMHs. 
Note that including or excluding
the membrane associated proteins 
has no effect on this statistical test, as the chance 
of a successful trial is zero (there is zero chance
to find a SNP in TMHs if there are none) \richel{GvdB: Also too complex for biologists. Explain in bit more words}.
The actually number of SNPs found in TMHs
was 3889, which is lower than the null hypothesis. 
The chance to find, within TMHs, this amount or less SNPs 
equals $1.6201 \cdot 10^{-9}$.
Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we conclude that there are significantly less SNPs
in TMHs as expected by chance.

@@

>>>>

The number of variations
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in \richel{282250} variations.

A variation needs not always be a SNP,
as dbSNP also catalogs other DNA alterations, such as, among others, insertions,
deletions and tandem repeats.
We select only the variations that result in a SNP for
a single amino acid substitution.

@@

In the final set of experiments, 
we addressed the question whether TMHs 
would be more evolutionary conserved than soluble proteins regions 
by comparing the occurrences of SNPs in the genome coding 
for TMHs and soluble protein regions within membrane proteins. 
We obtained 1,129 gene names associated with the phrase 'membrane protein',
which are genes for both membrane-associated proteins (MAPs, no TMH) and 
transmembrane proteins (TMPs, at least 1 TMH).
These genes are linked to 4,811 protein isoforms, 
of which 2,568 are predicted to be TMPs (hence 
2,243 proteins are predicted to be MAPs).
We obtained 38,882 SNPs that resulted in an
amino acids substitution, of which 21,738 were located 
in predicted TMPs.
Per protein, we calculated two percentages: 
(1) the percentage of the protein coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without MAPs,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs and soluble protein domains.
Our analysis shows that SNPs occur more often in soluble domains
than in TMHs.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

To statistically test whether SNPs occur 
just as likely in soluble domains as
in TMHs, we first calculated the number of SNPs that are expected to be
found in TMHs of TMPs by chance, $E(n_{SNP\_TMH})$,
equals the number of SNPs in TMHs, $n_{SNP} = 21,738$, 
multiplied by the fraction of the TMPs that are TMH, $f_{TMH} = 18.66\%$,
$E(n_{SNP\_TMH}) = n_{SNP} \cdot f_{TMH} \approx 4056$.

The actually number of SNPs found in TMHs
was $n_{SNP\_TMH} = 3,903$, which is lower than the expectation
of $E(n_{SNP\_TMH}) \approx 4056$.
To test whether this difference is significant,
we used a binomial Poisson distribution, where the number of trails,
$n$ equals the number of SNPs, which is 21,738. The probability of success
for TMP $i$, $p\_i$, is the percentage of each TMP that is TMH.
We calculated the probability of having 3,903 or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals \richel{recalculate}
% $1.6201 \cdot 10^{-9}$.

Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we conclude that there are significantly less SNPs
in TMHs as expected by chance.

@@

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_snps_per_gene_name_ncbi.png}
  \caption{
    Distribution of the number of SNPs per gene name in the NCBI database.
  }
  \label{fig:snps_per_gene_name_ncbi}
\end{figure}
